---
title: "Reproducible_Research_homework2"
author: "Yipu Ding"
output: html_document
date: "2025-03-11"
------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5)
```

# Loading and preprocessing the data

```{r, echo = TRUE}
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
Sys.setlocale("LC_TIME","English")

# Load the dataset
activity <- read_csv("C:/Users/aaact/Downloads/activity.csv")
```

# What is mean total number of steps taken per day?

## Calculate the total number of steps taken per day

```{r, echo = TRUE}
# Group by date and calculate total steps per day
total_steps_per_date <- activity %>% 
  group_by(date) %>% 
  summarize(total_steps = sum(steps, na.rm = TRUE))

print(total_steps_per_date)

```

## Make a histogram of the total number of steps taken each day

```{r, echo = TRUE}
# Create a bar plot (histogram) of total steps per day
ggplot(total_steps_per_date, aes(x = date, y = total_steps)) +
  geom_bar(
    stat = "identity", 
    fill = "#F69B85",  # Bar fill color
    color = "black",   # Bar border color (black and thin)
    alpha = 0.8,       # Transparency
    width = 0.8,       # Bar width
    size = 0.3         # Border thickness
  ) +
  labs(
    title = "Total Steps Taken Per Day",  
    x = "Date", 
    y = "Total Steps"
  ) +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.title = element_text(face = "bold"),  
    axis.line = element_line(color = "black", size = 0.5),  
    axis.ticks = element_line(color = "black", size = 0.5),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank() 
  ) +
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels with commas

```

## Calculate and report the mean and median of the total number of steps taken per day

```{r, echo = TRUE}
# Calculate mean and median steps per day
mean_median_steps <- activity %>% 
  group_by(date) %>% 
  summarize(
    total_steps_mean = mean(steps, na.rm = TRUE),
    total_steps_median = median(steps, na.rm = TRUE)
  )

print(mean_median_steps)

```

# What is the average daily activity pattern?

## Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
# Group by interval and calculate average steps
activity_time_series <- activity %>%
  group_by(interval) %>%
  summarize(average_steps = mean(steps, na.rm = TRUE))

# Create a time series plot
ggplot(activity_time_series, aes(x = interval, y = average_steps)) +
  geom_line(color = "#1F78B4", size = 0.8) +  
  labs(
    title = "Average Daily Activity Pattern",
    x = "5-Minute Interval", 
    y = "Average Number of Steps"
  ) +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.title = element_text(face = "bold"),  
    axis.line = element_line(color = "black", size = 0.5),  
    axis.ticks = element_line(color = "black", size = 0.5),  
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_blank() 
  )
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
# Find the interval with the maximum average steps
max_interval <- activity_time_series %>%
  filter(average_steps == max(average_steps, na.rm = TRUE))

print(max_interval)

```

The answer is 8:35, with 206 average steps.

# Imputing missing values

## Calculate and report the total number of missing values in the dataset

```{r}
# Count the number of rows with missing steps
missing_rows <- activity %>% filter(is.na(steps))
nrow(missing_rows)

```

The total number of missing values was 2304.

## Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated.

## Create a new dataset that is equal to the original dataset but with the missing data filled in.

Here, I will fill in the missing values using the mean of the 5-minute interval.

```{r}
# Fill missing values with the mean for that 5-minute interval
activity_filled <- activity %>%
  group_by(interval) %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))

print(activity_filled)

```

## Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
# Calculate total steps per day with imputed data
total_steps_per_date_filled <- activity_filled %>% 
  group_by(date) %>% 
  summarize(total_steps = sum(steps, na.rm = TRUE))

# Create a bar plot (histogram) with imputed data
ggplot(total_steps_per_date_filled, aes(x = date, y = total_steps)) +
  geom_bar(
    stat = "identity", 
    fill = "#F69B85",  # Bar fill color
    color = "black",   # Bar border color (black and thin)
    alpha = 0.8,       # Transparency
    width = 0.8,       # Bar width
    size = 0.3         # Border thickness
  ) +
  labs(
    title = "Total Steps Taken Per Day (with Imputed Data)", 
    x = "Date", 
    y = "Total Steps"
  ) +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.title = element_text(face = "bold"),  
    axis.line = element_line(color = "black", size = 0.5),  
    axis.ticks = element_line(color = "black", size = 0.5),  
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.background = element_blank()   # Remove background
  ) +
  scale_y_continuous(labels = scales::comma)  

# Calculate mean and median with imputed data
mean_median_steps_filled <- activity_filled %>% 
  group_by(date) %>% 
  summarize(
    total_steps_mean = mean(steps, na.rm = TRUE),
    total_steps_median = median(steps, na.rm = TRUE)
  )

print(mean_median_steps_filled)

```

# Are there differences in activity patterns between weekdays and weekends?

## Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
# Add a new column to distinguish weekdays and weekends
activity_weekday <- activity %>% 
  mutate(weekday = weekdays(date)) %>% 
  mutate(weekday_or_end = ifelse(weekday %in% c("Saturday", "Sunday"), "weekend", "weekday"))

print(activity_weekday)

```

## Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
# Calculate average steps per interval for weekdays and weekends
activity_time_series_weekday <- activity_weekday %>%
  group_by(interval, weekday_or_end) %>%
  summarize(average_steps = mean(steps, na.rm = TRUE))

# Create a panel plot with separate lines for weekdays and weekends
ggplot(activity_time_series_weekday, aes(x = interval, y = average_steps, color = weekday_or_end)) +
  geom_line(size = 0.8) +  
  facet_wrap(~weekday_or_end, ncol = 1, scales = "free_x") +  # Ensure each panel has its own x-axis
  labs(
    title = "Average Daily Activity Pattern (Weekday vs. Weekend)",
    x = "5-Minute Interval", 
    y = "Average Number of Steps",
    color = "Day Type"
  ) +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.title = element_text(face = "bold"),  
    axis.line = element_line(color = "black", size = 0.5),  
    axis.ticks = element_line(color = "black", size = 0.5),  
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.background = element_blank(),  # Remove background
    strip.text = element_text(face = "bold", size = 12)  # Style facet labels
  ) +
  scale_color_manual(values = c("weekday" = "#1F78B4", "weekend" = "#33A02C"))  # Custom line colors

```

There are differences in activity patterns between weekdays and weekends. Here are some observations:

1. Activity Intensity: On weekdays, activity intensity may be higher during specific times, such as early morning and late afternoon, likely due to commuting and work-related activities. On weekends, activity might be more evenly distributed, with potential peaks during mid-morning or early afternoon.

2. Activity Distribution: Weekend activities appear to be more spread out without distinct peaks, reflecting a more relaxed schedule and varied leisure activities.

3. Peak Times: Weekdays may show activity peaks in the morning and evening, while weekends could have peaks during mid-morning or early afternoon.

4. Activity Volume: Overall, activity levels on weekends might be higher during certain periods compared to weekdays, possibly because people have more free time for exercise or outdoor activities.

